Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-01-05 Manuscript/TEM_draft.tex
Updated BC section

@@

The effect of missing data on phylogenetic inferences has been widely studied \citep{wiensmissing2003,wiensmissing2006,wiensmissing2008,lemmonthe2009,rouresite-specific2011,sansomfossilization2013,pattinsonphylogeny2014,sansombias2014}. Missing molecular data has been seen by some authors as an issue because it can, in some parts of the tree, decrease phylogenetic signal (i.e. the evolutionary information contained within the matrix allowing the inference of topology and branch lengths), especially when using large matrices \citep{lemmonthe2009}. However, this may not be a major issue because phylogenetic signal is easily increased by: (i) including a "modest" number ($~$ % NC: This symbol doesn't show up on my PDF. Check.
50) of highly-covered genes (i.e. a number of genes that are available for all taxa; \citealt{rouresite-specific2011}); (ii) adding a greater number of taxa (especially slowly-evolving taxa or taxa close to the outgroup; \citealt{rouresite-specific2011}); and (iii) choosing more appropriate models of sequence evolution \citep{wiensmissing2006,wiensmissing2008,rouresite-specific2011}. Similarly, missing morphological data might be seen as either a major or minor issue for accurately inferring phylogenies depending on the study in question \citep{wiensmissing2003,sansomfossilization2013,pattinsonphylogeny2014}. Because soft-tissue characters are rarely preserved in the fossil record, missing data is mainly found in these characters, and is therefore not randomly distributed which can lead to biased placement of fossil taxa in phylogenies (e.g. \citealt{sansomfossilization2013}, but see \citealt{pattinsonphylogeny2014}). However, the phylogenetic signal is not related to the amount of missing data \textit{per se} but to the number of informative characters for each taxon, therefore missing data is less of an issue than the number of shared informative characters \citep{kearneyfragmentary2002,wiensmissing2003,pattinsonphylogeny2014}.
% NC: I'm a little dubious of what this paragraph is trying to get at in places. We should discuss at some point. TG: my idea was to do the "literature-review" part of the intro: missing data is already widely studied (answer to the comment: "hun... We already knew that") but it has never been studied in TEM framework. And that's kind of the point we're getting toward with our work: we don't really care about the effect of missing data per se (i.e. what's the threshold where it starts fucking up the phylogeny) but more the fact that Bayesian fails.
% NC: OK fair enough. I think though if this is too long we can probably cut this paragraph and put a bit of it into the paragraph below.

Although missing data does not appear be a major problem in molecular and morphological matrices separately \citep{wiensmissing2003,wiensmissing2006,wiensmissing2008,rouresite-specific2011,pattinsonphylogeny2014}, it may become more of an issue in Total Evidence matrices containing both molecular and morphological data for living and fossil species. This may be particularly problematic as fossil taxa (generally) do not have molecular data, resulting in a large section of missing data in Total Evidence matrices. Until now, no attempt has been made to study the impact of this issue on phylogenetic inference from Total Evidence methods. 
% NC: THIS SECTION NEEDS WORK!
In this study, we focus on the effect of missing data on the ability to recover a correct topology, 
% NC: I don't get the logic of this part 
% because topology is the first and most straightforward aspect reflecting phylogenetic signal (i.e. topological changes are discrete opposed to branch length changes are continuous). 
% NC: Need a better justification of WHY topology is interesting.
Although correctly inferring branch lengths is also important, we do not consider the inference of branch lengths here.
% NC: WHY??? Difficulties with simulations? Topology is more important? WHY?
Also, the effect of Total Evidence method has not been formally assessed in previous studies using fixed topology \citep{ronquista2012,schragocombining2013,slaterphylogenetic2013,beckancient2014}.
% NC: OK this last sentence could be worked in earlier on to explain WHY topology not branch lengths.

@@

\begin{figure}
\centering
    \includegraphics[keepaspectratio=true]{Figures/In_main/Simulations_outline-BW.pdf}
\caption{Protocol outline.
(1) We randomly generated a birth-death tree (the "true" tree) and used it to infer a matrix with no missing data (the "complete" matrix).
(2) We removed data from the morphological part of the "complete" matrix resulting in 125 "missing-data" matrices.
(3) We built phylogenetic trees from each matrix using both Maximum Likelihood and Bayesian methods.
(4) We compared the "missing-data" trees to the "best" tree.
We repeated these four steps 50 times.}
\label{Fig_Outline}
\end{figure}
% NC: Can you centre these around the boxes? and make the comparing topologies arrow touch the two triangles? I'll show you what I mean. 
% NC: Also check the figure is in a format and of correct quality for Syst Biol.

%---------------------------------------------
%       Generating the matrix
%---------------------------------------------

\subsection{Generating the matrix} % NC: Can you number these subsections like they are in the steps above?
First we randomly generated a "true" tree of 50 taxa in R v. 3.0.2 \citep{R302} using the package diversitree v. 0.9-6 \citep{fitzjohndiversitree2012}. We generated the tree using a birth death process by sampling speciation ($\lambda$) and extinction ($\mu$) rates from a uniform distribution but maintaining $\lambda$ $>$ $\mu$ \citep{paradistime-dependent2011}. We implemented a rejection sampling algorithm to select only trees with 25 living and 25 fossil taxa to ensure that we had enough taxa of each type for our missing data simulations to work. We then added an outgroup to the tree, using the mean branch length of the tree to separate the outgroup from the rest of the taxa, and with the branch length leading to the outgroup set as the sum of the mean branch length and the longest root-to-tip length of the tree.

@@

%---------------------------------------------
%       Removing data
%---------------------------------------------

\subsection{Removing data}
\label{Removing_data}
We modified the "complete" matrix to get matrices with missing data by randomly replacing data with "?" in the morphological part of the matrices according to the following parameters:% (Fig. ~\ref{Fig_RemoveData}): % PUT IN APPENDIX

@@

%---------------------------------------------
%       Building phylogenies
%---------------------------------------------

\subsection{Building phylogenies}
From the resulting matrices we generated two types of trees: the "best" tree inferred from the "complete" matrix and the "missing-data" trees inferred from the 125 matrices with various amounts of missing data. The "true" tree was used to generate the "complete" matrix and reflects the "true" evolutionary history in our simulations. The "best" tree, on the other hand, is the best tree we can build using state-of-the-art phylogenetic methods. In real world situations, the "true" tree is never available to us because we cannot know the true evolutionary history of a clade (except in very rare circumstances, e.g. \citealt{rozen2005}). Therefore, here we focus on comparing the trees inferred from the matrices with missing data to the "best" tree, rather than the "true" tree, as the "best" tree is generally what we have to work with.

@@

To measure the phylogenetic signal of our simulations, we first ran a fast bootstrap analysis with 500 replicates on the "complete" matrix. We removed all the simulations with a median bootstrap support lower than 50 as a proxy for weak phylogenetic signal \citep{zanderminimal2004}. We repeated this selection until we obtained 50 sets of simulations (i.e. 50 "complete" and 50 x 125 "missing-data" matrices) with a relative good phylogenetic signal (median bootstrap $>$ 50). On these selected simulations, we used the fast bootstrap algorithm and performed 1000 bootstraps for each tree inference to assess topological support \citep{pattengale2010many}.
%The bootstrap algorithm used in RAxML is the Lazy Sub-tree Rearrangement (LSR) which consists of pruning one sub-tree from the tree and subsequently reinserting it to all neighbouring branches \citep{stamatakisa2008}. Sub-tree Pruning and Reinserting methods (SPR) have been demonstrated to be better than other methods (e.g. Nearest Neighbouring Interchange - NNI) for recovering good bootstrap values \citep{salamin2003}. %TG: The explanations of the algorithm part for the BS was proposed by Trevor I think it's useless
% NC: perhaps somewhere in brackets just mention it's LSR bootstrapping. No need for the rest I agree.
Using these parameters took $~$8 % NC: again tilde isn't appearing on my PDF 
CPU years to build 50 sets of 125 bootstrapped Maximum Likelihood trees (2.30GHz clock speed nodes).

@@

Each Bayesian tree was estimated using two runs of four chains each for a maximum of 50$\times$$10^6$ generations.
% NC: Do you mean 50 x 10^6? or 5 x 10^7? It's standard to use 10^x not 1^x
 We used the average standard deviation of split frequencies (ASDS) as a proxy to estimate the convergence of the chains and used a stop rule when the ASDS went below 0.01 \citep{Ronquist2012mrbayes}. The effective sample size (ESS) was also checked on a random sub-sample of runs in each simulation to ensure that ESS $>>$ 200 \citep{drummond2006ess}. For each run, we removed 25\% of the iterations as burn-in. We used the following priors for each tree (see \hyperref[SupplementaryMaterial]{Supplementary Material S1}
 % NC: For what? What's in the SI?
 ): (i) the "true" tree’s topology as a starting tree (with a starting value for each branch length of one), (ii) an exponential prior on the shape of the gamma distribution of $\alpha$ = 0.5 for both partitions, and (iii) a transition/transversion ratio prior of two sampled from a strong beta distribution ($\beta$(80,40)).

We used these priors to speed up the Bayesian estimation process. These priors biased the way the Bayesian process calculated branch lengths by giving non-random starting points and boundaries for parameter estimation however, here we are focusing on the effect of missing data on tree topology and not branch lengths. Even using these priors, it took $~$ % NC: Another tilde issue
140 CPU years to build 50 sets of 125 Bayesian trees (2.30GHz clock speed nodes).

%---------------------------------------------
%       Comparing Topologies
%---------------------------------------------

\subsection{Comparing topologies}
We compared the topology of the "missing-data" trees to the "best" tree to measure the effect of the three parameters $M_{L}$, $M_{F}$ and $M_{C}$ on tree topology. We used the Robinson-Foulds distance \citep{RF1981} to assess the amount of conserved clade positions and the Triplets distance \citep{dobson1975triplets} to assess the amount of wildcard taxa \citep{kearneyfragmentary2002}. We then used the Normalized Tree Similarity index \citep{Bogdanowicz2012} to generalize our results for any \textit{n} number of taxa. These metrics are described in detail below.

\subsubsection{Robinson-Foulds distance}
The Robinson-Foulds distance \citep{RF1981}, or "path difference", measures the number of shared clades across two trees. The metric reflects the distance between the distributions of tips among clades in the two trees (\citealt{RF1981}; see \hyperref[SupplementaryMaterial]{Supplementary Material S2}
% NC: For what? What's in the SI?
). This metric is bounded between one, when the two trees are identical, and $n-2$ (for two trees with $n$ taxa) when there is no shared clade in the two trees. This metric is sensitive to even minor changes in clade conservation: if the trees are composed of two clades of three taxa (\textit{(((a,b),c),((d,e),f))}), the swapping of any two taxa will lead to a maximal score of the Robinson-Foulds distance indicating poor tree similarity.

\subsubsection{Triplets distance}
The Triplets distance \citep{dobson1975triplets} measures the number of sub-trees made up of three taxa that differ between two trees (\citealt{critchlowthe1996}; see \hyperref[SupplementaryMaterial]{Supplementary Material S2}
% NC: For what? What's in the SI?
). This metric measures the position of each taxon and clade in relation to its closest neighbours. It is bounded between 0 when the two trees are identical and $\binom{n}{4}$ (for two trees with $n$ taxa) when there is no shared taxon/clade position in the two trees. Therefore this metric sensitive to the conservation of wildcard taxa, i.e. taxa that frequently change position in different trees.
% NC: You may have a better definition of wildcard taxa than mine :)

\subsubsection{Normalized Tree Similarity}
The Normalized Tree Similarity index, $NTS_m$ (\citealp{Bogdanowicz2012}) allows us to compare the two metrics above for any number of taxa, by scaling the value of the metric $m$ (either Robinson-Foulds or Triplets distance in our study) to the expected value of the metric $m$ when comparing two random trees (see \hyperref[SupplementaryMaterial]{Supplementary Material S2}
% NC: Why? For more details? Simulations etc
). When $NTS_m$=1, the two trees are identical, when $NTS_m$=0 the trees are no more different than expected when comparing two random trees, and when $NTS_m$$<$0, the difference between the two trees is greater than when comparing two random trees. In our study we used the $NTS_m$ index as a proxy for topological similarity: a high score of this index (i.e. towards one) means that the topology is highly conserved between the two trees; conversely, a low score of this index (i.e. towards 0) means that the topological difference between the two trees is as great as expected when comparing two random trees. Note that the two tree comparison metrics represent different aspects of topological similarity and do not perform equally well (see \hyperref[metrics_discussion]{discussion below} and \citealt{kuhnerpractical2014}).

% NC: If I understand right, you use RF normalised and TD normalised. NOT RF, TD and NTS. So having these as three separate sections seems odd. Instead I'd write RF section, then explain that it is not comparable across trees with different numbers of taxa so you normalise it as suggested in Bogdanowicz's NTS method. Then describe TD and explain that you normalise it as discussed above for RF. NTS isn't really a metric, just a simple normalisation procedure.

\subsubsection{Tree comparisons} % NC: OK I'm lost in this section! We should discuss then think about how to express this in an understandable fashion.
\label{tree_comparisons}
For the Maximum Likelihood and Bayesian consensus
% NC: is this actually a maximum clade credibility tree
trees we performed pairwise comparisons between the "best" tree and each "missing-data" tree using both the Robinson-Foulds and Triplets metrics with the TreeCmp java script \citep{Bogdanowicz2012}. For each metric, we then normalized the value using the Normalized Tree Similarity scaled by the mean value of 1000 pairwise random tree comparisons for the metric in question and n = 51 taxa (see \hyperref[SupplementaryMaterial]{Supplementary Material Section 2}). % TG: Maybe remove supplementary here
% NC: Yeah probably. You could also fit this into the section above. Again NTS isn't really a metric in my opinion...
%We compared each "missing-data" tree with the "best" tree for each of our 50 simulation runs resulting in 50 comparisons for each "missing-data" tree. We calculated the mode and the 50\% and 95\% confidence intervals from the resulting distribution using the hdrcde R package v3.1 \citep{hdrcde}. TG: explained later on

To take into account the uncertainty of tree inference in both Maximum Likelihood and Bayesian trees, % NC: instead of using the single ML or B MCC tree as above,
we extracted 1000 random bootstrapped trees from the Maximum Likelihood analysis and 1000 from the posterior tree distribution of the Bayesian analysis. % for the best trees and then did the same for the 125 missing data trees (resulting in 125 x 1000 missing data trees). 
For a given set of 1000 missing data trees, we sampled one missing data tree and one best tree at random and compared them using both the Robinson-Foulds and Triplets metrics as described above. We repeated this 1000 times for each set of missing data trees resulting in 125 x 1000 values for each metric.

We repeated all of the above for each of the 50 chains. We then calculated the mode and the 50\% and 95\% confidence intervals from the resulting distribution using the hdrcde R package v. 3.1 \citep{hdrcde}. %Finally we tested for significant differences among the different distributions of trees using non-parametric group and pairwise comparisons by using respectively Kruskal-Wallis and Nemenyi-Damico-Wolfe-Dunn test (CITE R PACKAGE)\citet{ruxtontime2008}. TG: finally we didn't (Andrew's advice - see Battacharyya)

% NC: New subheading (summarising)

% NC: Finally after doing all theses stupid comparisons and simulations we tested the effects of our missing data parameters (ML) on our ability to recover the correct topology. 
To do this we measured the similarity among the different distributions % NC: OF WHAT!!! tree metrics for all of the analyses above
using the Bhattacharyya Coefficient \citep{Bhattacharyya}. The Bhattacharyya Coefficient is the probability of overlap between two distributions, ranging from 0 to 1 (\citealt{Bhattacharyya} - see \hyperref[SupplementaryMaterial]{Supplementary Material Section 2} % FOR WHAT!!!).

>>>>

%LaTeX template : http://systbio.org/files/SB_LaTeX_Template_txt_extension.txt
%Author instructions: http://www.oxfordjournals.org/our_journals/sysbio/for_authors/ms_preparation.html

@@

%TG: I completely removed this paragraph
%The effect of missing data on phylogenetic inferences has been widely studied \citep{wiensmissing2003,wiensmissing2006,wiensmissing2008,lemmonthe2009,rouresite-specific2011,sansomfossilization2013,pattinsonphylogeny2014,sansombias2014}. Missing molecular data has been seen by some authors as an issue because it can, in some parts of the tree, decrease phylogenetic signal (i.e. the evolutionary information contained within the matrix allowing the inference of topology and branch lengths), especially when using large matrices \citep{lemmonthe2009}. However, this may not be a major issue because phylogenetic signal is easily increased by: (i) including a "modest" number (\texttildelow % NC: This symbol doesn't show up on my PDF. Check.
%50) of highly-covered genes (i.e. a number of genes that are available for all taxa; \citealt{rouresite-specific2011}); (ii) adding a greater number of taxa (especially slowly-evolving taxa or taxa close to the outgroup; \citealt{rouresite-specific2011}); and (iii) choosing more appropriate models of sequence evolution \citep{wiensmissing2006,wiensmissing2008,rouresite-specific2011}. Similarly, missing morphological data might be seen as either a major or minor issue for accurately inferring phylogenies depending on the study in question \citep{wiensmissing2003,sansomfossilization2013,pattinsonphylogeny2014}. Because soft-tissue characters are rarely preserved in the fossil record, missing data is mainly found in these characters, and is therefore not randomly distributed which can lead to biased placement of fossil taxa in phylogenies (e.g. \citealt{sansomfossilization2013}, but see \citealt{pattinsonphylogeny2014}). However, the phylogenetic signal is not related to the amount of missing data \textit{per se} but to the number of informative characters for each taxon, therefore missing data is less of an issue than the number of shared informative characters \citep{kearneyfragmentary2002,wiensmissing2003,pattinsonphylogeny2014}.
% NC: I'm a little dubious of what this paragraph is trying to get at in places. We should discuss at some point. TG: my idea was to do the "literature-review" part of the intro: missing data is already widely studied (answer to the comment: "hun... We already knew that") but it has never been studied in TEM framework. And that's kind of the point we're getting toward with our work: we don't really care about the effect of missing data per se (i.e. what's the threshold where it starts fucking up the phylogeny) but more the fact that Bayesian fails.
% NC: OK fair enough. I think though if this is too long we can probably cut this paragraph and put a bit of it into the paragraph below.

Although missing data does not appear be a major problem in molecular and morphological matrices separately (as long as enough data overlaps in each case) \citep{wiensmissing2003,wiensmissing2006,wiensmissing2008,rouresite-specific2011,pattinsonphylogeny2014}, it may become more of an issue in Total Evidence matrices containing both molecular and morphological data for living and fossil taxa. This may be particularly problematic as fossil taxa (generally) do not have molecular data, resulting in a large section of missing data in Total Evidence matrices. Until now, no attempt has been made to study the impact of this issue on phylogenetic inference using Total Evidence methods.
% NC: THIS SECTION NEEDS WORK!
It is expected that missing in this framework will affect the ability to recover the correct topology as well as the correct branch length estimation \citep{wiensmissing2003,wiensmissing2006,wiensmissing2008,lemmonthe2009,rouresite-specific2011,sansomfossilization2013,pattinsonphylogeny2014,sansombias2014}. In this study, we focus only on the effect of missing data on the ability to recover a correct topology because it is a crucial aspect of a phylogeny in many macroevolutionary studies \citep[e.g.][]{manosphylogeny2007,meredithimpacts2011,jetzthe2012}. %TG or at some more specific example instead of the cite: "In this study, we focus only on the effect of missing data on the ability to recover a correct topology because it is a crucial aspect of a phylogeny in many macroevolutionary studies (e.g. for taxonomical focus \citep{meredithimpacts2011,jetzthe2012}, or for studying evolutionary trends %CITE?\citep[e.g.][]{})"?
Even though branch length is another primordial aspect in phylogenies (namely for timing extinction and/or speciation events \citep[e.g.][]{ronquista2012}), previous studies have empirically assessed the effect of the Total Evidence method on branch length variation \citep{ronquista2012,schragocombining2013,slaterphylogenetic2013,beckancient2014} %TG: add:"by using a fixed topology"?
Moreover, variation in branch length are more complex to simulate and to measure when topology varies.
% NC: I don't get the logic of this part 
% because topology is the first and most straightforward aspect reflecting phylogenetic signal (i.e. topological changes are discrete opposed to branch length changes are continuous). 
% NC: Need a better justification of WHY topology is interesting.

%Although correctly inferring branch lengths is also important, we do not consider the inference of branch lengths here.
% NC: WHY??? Difficulties with simulations? Topology is more important? WHY?
%Also, the effect of Total Evidence method has not been formally assessed in previous studies using fixed topology \citep{ronquista2012,schragocombining2013,slaterphylogenetic2013,beckancient2014}.
% NC: OK this last sentence could be worked in earlier on to explain WHY topology not branch lengths.

@@

\begin{figure}
\centering
    \includegraphics[keepaspectratio=true]{Figures/In_main/Simulations_outline-BW.pdf}
\caption{Protocol outline.
(1) We randomly generated a birth-death tree (the "true" tree) and used it to infer a matrix with no missing data (the "complete" matrix).
(2) We removed data from the morphological part of the "complete" matrix resulting in 125 "missing-data" matrices.
(3) We built phylogenetic trees from each matrix using both Maximum Likelihood and Bayesian methods.
(4) We compared the "missing-data" trees to the "best" tree.
We repeated these four steps 50 times.}
\label{Fig_Outline}
\end{figure}
% NC: Can you centre these around the boxes? and make the comparing topologies arrow touch the two triangles? I'll show you what I mean. 
% TG: I have different versions for the figure available (you can see them opening it with inkscape I think - else I'll show you).
% NC: Also check the figure is in a format and of correct quality for Syst Biol.

%---------------------------------------------
%       1-Generating the matrix
%---------------------------------------------

\subsection{1-Generating the matrix} % NC: Can you number these subsections like they are in the steps above? % TG: 
First we randomly generated a "true" tree of 50 taxa in R v. 3.0.2 \citep{R302} using the package diversitree v. 0.9-6 \citep{fitzjohndiversitree2012}. We generated the tree using a birth death process by sampling speciation ($\lambda$) and extinction ($\mu$) rates from a uniform distribution but maintaining $\lambda$ $>$ $\mu$ \citep{paradistime-dependent2011}. We implemented a rejection sampling algorithm to select only trees with 25 living and 25 fossil taxa to ensure that we had enough taxa of each type for our missing data simulations to work. We then added an outgroup to the tree, using the mean branch length of the tree to separate the outgroup from the rest of the taxa, and with the branch length leading to the outgroup set as the sum of the mean branch length and the longest root-to-tip length of the tree.

@@

%---------------------------------------------
%       2-Removing data
%---------------------------------------------

\subsection{2-Removing data}
\label{Removing_data}
We modified the "complete" matrix to get matrices with missing data by randomly replacing data with "?" in the morphological part of the matrices according to the following parameters:

@@

%---------------------------------------------
%       3-Building phylogenies
%---------------------------------------------

\subsection{3-Building phylogenies}
From the resulting matrices we generated two types of trees: the "best" tree inferred from the "complete" matrix and the "missing-data" trees inferred from the 125 matrices with various amounts of missing data. The "true" tree was used to generate the "complete" matrix and reflects the "true" evolutionary history in our simulations. The "best" tree, on the other hand, is the best tree we can build using state-of-the-art phylogenetic methods. In real world situations, the "true" tree is never available to us because we cannot know the true evolutionary history of a clade (except in very rare circumstances, e.g. \citealt{rozen2005}). Therefore, here we focus on comparing the trees inferred from the matrices with missing data to the "best" tree, rather than the "true" tree, as the "best" tree is generally what we have to work with.

@@

To measure the phylogenetic signal of our simulations, we first ran a fast bootstrap analysis with 500 replicates on the "complete" matrix. We removed all the simulations with a median bootstrap support lower than 50 as a proxy for weak phylogenetic signal \citep{zanderminimal2004}. We repeated this selection until we obtained 50 sets of simulations (i.e. 50 "complete" and 50 x 125 "missing-data" matrices) with a relative good phylogenetic signal (median bootstrap $>$ 50). On these selected simulations, we used the fast bootstrap algorithm and performed 1000 bootstraps for each tree inference to assess topological support \citep{pattengale2010many}.
%The bootstrap algorithm used in RAxML is the Lazy Sub-tree Rearrangement (LSR) which consists of pruning one sub-tree from the tree and subsequently reinserting it to all neighbouring branches \citep{stamatakisa2008}. Sub-tree Pruning and Reinserting methods (SPR) have been demonstrated to be better than other methods (e.g. Nearest Neighbouring Interchange - NNI) for recovering good bootstrap values \citep{salamin2003}. %TG: The explanations of the algorithm part for the BS was proposed by Trevor I think it's useless
% NC: perhaps somewhere in brackets just mention it's LSR bootstrapping. No need for the rest I agree.
Using these parameters took \texttildelow 8 CPU years to build 50 sets of 125 bootstrapped Maximum Likelihood trees (2.30GHz clock speed nodes).

@@

Each Bayesian tree was estimated using two runs of four chains each for a maximum of 5$\times$$10^7$ generations. We used the following priors for each tree: (i) the "true" tree’s topology as a starting tree (with a starting value for each branch length of one), (ii) an exponential prior on the shape of the gamma distribution of $\alpha$ = 0.5 for both partitions, and (iii) a transition/transversion ratio prior of two sampled from a strong beta distribution ($\beta$(80,40)). We used these priors to speed up the Bayesian estimation process. These priors biased the way the Bayesian process calculated branch lengths by giving non-random starting points and boundaries for parameter estimation however, here we are focusing on the effect of missing data on tree topology and not branch lengths. Even using these priors, it took \texttildelow 140 CPU years to build 50 sets of 125 Bayesian trees (2.30GHz clock speed nodes). The detailed MrBayes parameters are available in the \hyperref[SupplementaryMaterial]{Supplementary Material S1}.

We used the average standard deviation of split frequencies (ASDS) as a proxy to estimate the convergence of the chains and used a stop rule when the ASDS went below 0.01 \citep{Ronquist2012mrbayes}. The effective sample size (ESS) was also checked on a random sub-sample of runs in each simulation to ensure that ESS $>>$ 200 \citep{drummond2006ess}. Finally we built the Bayesian consensus tree from the combined chains excluding the 25\% first iterations (burn-in) with a strict majority rule \citep{Ronquist2012mrbayes}.
%---------------------------------------------
%       4-Comparing Topologies
%---------------------------------------------

\subsection{4-Comparing topologies}
We compared the topology of the "missing-data" trees to the "best" tree to measure the effect of the three parameters $M_{L}$, $M_{F}$ and $M_{C}$ on tree topology. We used the Robinson-Foulds distance \citep{RF1981} to assess the amount of conserved clade positions and the Triplets distance \citep{dobson1975triplets} to assess the amount of wildcard taxa  (i.e. taxa that frequently change position in different trees - \citealt{kearneyfragmentary2002}). We normalised both metrics using the Normalized Tree Similarity index \citep{Bogdanowicz2012} to generalize our results for any \textit{n} number of taxa. These metrics are described in detail below.

\subsubsection{Robinson-Foulds distance}
The Robinson-Foulds distance \citep{RF1981}, or "path difference", measures the number of shared clades across two trees. The metric reflects the distance between the distributions of tips among clades in the two trees (\citealt{RF1981}; see \hyperref[SupplementaryMaterial]{Supplementary Material S2} for calculation details). This metric is bounded between one, when the two trees are identical, and $n-2$ (for two trees with $n$ taxa) when there is no shared clade in the two trees. This metric is sensitive to minor changes in clade conservation: if the trees are composed of two clades of three taxa (\textit{(((a,b),c),((d,e),f))}), the swapping of any two taxa will lead to a maximal score of the Robinson-Foulds distance indicating poor tree similarity. We normalised this metric using the Normalized Tree Similarity index \citep{Bogdanowicz2012}. This index allows us to compare the metric for any number of taxa by scaling it with the expected value of the metric when comparing two random trees (see \hyperref[SupplementaryMaterial]{Supplementary Material S2} for the calculation details). Therefore, when the Normalised Robinson-Foulds distance between two trees is equal to one, the trees are identical; if the Normalised Robison-Foulds distance is equal to zero, the trees are no more different than expected by chance; finally if the Noramlised Robison-Foulds distance is lower than zero, the trees are more different than expected by chance.

\subsubsection{Triplets distance}
The Triplets distance \citep{dobson1975triplets} measures the number of sub-trees made up of three taxa that differ between two trees (\citealt{critchlowthe1996}; see \hyperref[SupplementaryMaterial]{Supplementary Material S2} for calculation details). This metric measures the position of each taxon and clade in relation to its closest neighbours. It is bounded between 0 when the two trees are identical and $\binom{n}{4}$ (for two trees with $n$ taxa) when there is no shared taxon/clade position in the two trees. Therefore this metric sensitive to the conservation of wildcard taxa. We normalised this metric in the same way as we did for the Robison-Foulds distance resulting in the Normalised Triplets distance.

%\subsubsection{Normalized Tree Similarity}
%The Normalized Tree Similarity index, $NTS_m$ (\citealp{Bogdanowicz2012}) allows us to compare the two metrics above for any number of taxa, by scaling the value of the metric $m$ (either Robinson-Foulds or Triplets distance in our study) to the expected value of the metric $m$ when comparing two random trees (see \hyperref[SupplementaryMaterial]{Supplementary Material S2}). When $NTS_m$=1, the two trees are identical, when $NTS_m$=0 the trees are no more different than expected when comparing two random trees, and when $NTS_m$$<$0, the difference between the two trees is greater than when comparing two random trees. In our study we used the $NTS_m$ index as a proxy for topological similarity: a high score of this index (i.e. towards one) means that the topology is highly conserved between the two trees; conversely, a low score of this index (i.e. towards 0) means that the topological difference between the two trees is as great as expected when comparing two random trees. Note that the two tree comparison metrics represent different aspects of topological similarity and do not perform equally well (see \hyperref[metrics_discussion]{discussion below} and \citealt{kuhnerpractical2014}).

\subsubsection{Paired trees comparisons} % NC: OK I'm lost in this section! We should discuss then think about how to express this in an understandable fashion. TG: or no subheading at all?
\label{tree_comparisons}
For the Maximum Likelihood and Bayesian consensus
% NC: is this actually a maximum clade credibility tree - TG: better stick with consensus as described above?
trees we performed pairwise comparisons between the "best" tree and each "missing-data" tree using both the Normalised Robinson-Foulds and Normalised Triplets metrics with the TreeCmp java script \citep{Bogdanowicz2012} resulting each time in 125 Normalised Robison-Foulds distances and 125 Normalised Triplets distance. %For each metric, we then normalized the value using the Normalized Tree Similarity scaled by the mean value of 1000 pairwise random tree comparisons for the metric in question and n = 51 taxa (see \hyperref[SupplementaryMaterial]{Supplementary Material Section 2}). % TG: That is now described in the RF distance subsection just above
%We compared each "missing-data" tree with the "best" tree for each of our 50 simulation runs resulting in 50 comparisons for each "missing-data" tree. We calculated the mode and the 50\% and 95\% confidence intervals from the resulting distribution using the hdrcde R package v3.1 \citep{hdrcde}. TG: explained later on
Also, to take into account the uncertainty of tree inference in both method, instead of using the single Maximum Likelihood and Bayesian consensus trees as above, we extracted 1000 random bootstrapped trees from the Maximum Likelihood analysis and 1000 from the posterior tree distribution of the Bayesian analysis for the "best" trees and then did the same for the 125 "missing data" trees (resulting in 1000 "best" trees and 125$\times$1000 "missing data" trees). 
For a given set of 1000 "missing data" trees and the 1000 "best" trees, we sampled one "missing data" tree and one "best" tree at random and compared them using both the Normalised Robinson-Foulds and Normalised Triplets metrics as described above. We repeated this 1000 times for each set of "missing data" trees resulting in 125$\times$1000 values for each metric. We repeated all the paired trees comparisons described above for each of the 50 simulations. We then calculated the mode and the 50\% and 95\% confidence intervals from the resulting distribution using the hdrcde R package v. 3.1 \citep{hdrcde}. %Finally we tested for significant differences among the different distributions of trees using non-parametric group and pairwise comparisons by using respectively Kruskal-Wallis and Nemenyi-Damico-Wolfe-Dunn test (CITE R PACKAGE)\citet{ruxtontime2008}. TG: finally we didn't (Andrew's advice - see Battacharyya)

\subsubsection{Summarizing the comparisons}
Finally after doing all theses %stupid
50 simulations and comparisons within, we tested the effects of our missing data parameters ($M_{L}$, $M_{F}$ and $M_{C}$) on the ability to recover the correct topology in a Total Evidence framework.
% TG: the following part might be useless
To
To do this we measured the similarity among the different distributions % NC: OF WHAT!!! tree metrics for all of the analyses above
using the Bhattacharyya Coefficient \citep{Bhattacharyya}. The Bhattacharyya Coefficient is the probability of overlap between two distributions, ranging from 0 to 1 (\citealt{Bhattacharyya} - see \hyperref[SupplementaryMaterial]{Supplementary Material Section 2} % FOR WHAT!!!).
